---
title: "Getting Started with FinancialFunctions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with FinancialFunctions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  digits   = 4
)
library(FinancialFunctions)
```

```{css, echo = FALSE}
h2 { border-bottom: 2px solid #0f3460; padding-bottom: 6px; color: #0f3460; }
h3 { color: #16213e; margin-top: 1.5rem; }
.fn-card {
  background: #f8faff;
  border-left: 4px solid #0f3460;
  padding: 0.8rem 1.2rem;
  border-radius: 0 6px 6px 0;
  margin: 1.5rem 0 0.3rem 0;
  font-family: monospace;
  font-size: 1.05em;
  color: #b7410e;
  font-weight: bold;
}
.purpose {
  color: #444;
  font-style: italic;
  margin: 0.3rem 0 0.8rem 0;
  font-family: sans-serif;
  font-weight: normal;
  font-size: 0.97em;
}
table { width: 100%; border-collapse: collapse; margin: 1rem 0; }
th { background: #0f3460; color: white; padding: 8px 12px; text-align: left; }
td { padding: 7px 12px; border-bottom: 1px solid #e0e0e0; }
tr:nth-child(even) { background: #f8faff; }
```

## Overview

**FinancialFunctions** provides 55 Excel-compatible financial functions
implemented in Rust for high-performance calculations directly inside R.
Every function:

- Produces results **identical to Excel** for the same inputs
- Returns **`NA`** on invalid or out-of-range input — never crashes
- Works on **scalar values** and inside `dplyr::mutate()` pipelines

```{r}
library(FinancialFunctions)
```

The 55 functions are organised into six groups:

| Group | Description | Count |
|-------|-------------|------:|
| 💰 Annuity | Loan payments, present/future values, interest rates | 10 |
| 📈 Cash Flows | NPV, IRR and irregular-date variants | 5 |
| 🏭 Depreciation | Asset write-down under various accounting methods | 7 |
| 🗓️ Coupon Dates | Bond coupon period date arithmetic | 6 |
| 🏦 Bonds & Securities | Pricing, yields, duration, accrued interest | 13 |
| 🔧 Miscellaneous | Rate conversions, T-bills, dollar notation, growth | 14 |

---

## 💰 Annuity Functions

Annuity functions answer the core time-value-of-money questions in corporate
finance, personal lending, and savings planning. They model a stream of fixed
payments at a fixed interest rate over a fixed number of periods.

---

<div class="fn-card">fv(rate, nper, pmt, pv, pmt_at_beginning)</div>
<div class="purpose">Future Value — how much will a series of payments and/or
a lump sum grow to over time at a given interest rate? Used to project
the terminal value of a savings plan, pension pot, or investment portfolio.</div>

```{r}
# Saving $200/month for 10 months at 6% annual, starting with $500
# Excel: FV(0.06/12, 10, -200, -500, TRUE) = 2581.40
fv(0.06/12, 10, -200, -500, TRUE)

# Monthly savings of $1,000 for 12 months at 12% annual
fv(0.12/12, 12, -1000, 0, FALSE)

# Guard: zero rate — FV is simply the sum of all payments
fv(0.0, 10, -100, -500, FALSE)
```

---

<div class="fn-card">pv(rate, nper, pmt, fv, pmt_at_beginning)</div>
<div class="purpose">Present Value — what is a future stream of payments worth
in today's money? Used to value annuities, mortgages, pensions, and any
fixed-income obligation.</div>

```{r}
# What is a pension paying $500/month for 20 years worth today at 8%?
# Excel: PV(0.08/12, 20*12, 500, 0, FALSE) = -59777.15
pv(0.08/12, 240, 500, 0, FALSE)
```

---

<div class="fn-card">pmt(rate, nper, pv, fv, pmt_at_beginning)</div>
<div class="purpose">Payment — what fixed periodic payment is required to
repay a loan or reach a savings target? The workhorse of mortgage, car loan,
and savings plan calculations.</div>

```{r}
# Monthly repayment on a $10,000 loan at 8% over 10 months
# Excel: PMT(0.08/12, 10, 10000, 0, FALSE) = -1037.03
pmt(0.08/12, 10, 10000, 0, FALSE)

# Guard: zero rate — payment = loan / periods
pmt(0.0, 12, 1200, 0, FALSE)
```

---

<div class="fn-card">ipmt(rate, per, nper, pv, fv, pmt_at_beginning)</div>
<div class="purpose">Interest Payment — of a periodic payment, how much goes
toward interest? Essential for building loan amortisation tables and for
reporting mortgage interest for tax purposes. Returns NA if per is outside
[1, nper].</div>

```{r}
# Interest in period 1 of an $8,000 loan at 10% over 3 months
# Excel: IPMT(0.10/12, 1, 3, 8000, 0, FALSE) = -66.67
ipmt(0.10/12, 1, 3, 8000, 0, FALSE)

# Guard: period = 0 returns NA
ipmt(0.10/12, 0, 12, 10000, 0, FALSE)

# Guard: period > nper returns NA
ipmt(0.10/12, 99, 12, 10000, 0, FALSE)
```

---

<div class="fn-card">ppmt(rate, per, nper, pv, fv, pmt_at_beginning)</div>
<div class="purpose">Principal Payment — of a periodic payment, how much
reduces the outstanding loan balance? Together with ipmt(), this builds a
complete amortisation schedule showing the split between interest and
principal for every period.</div>

```{r}
# Principal in period 1 of a $2,000 loan at 10% over 24 months
# Excel: PPMT(0.10/12, 1, 24, 2000, 0, FALSE) = -75.62
ppmt(0.10/12, 1, 24, 2000, 0, FALSE)
```

**Full amortisation schedule — first 6 months of a $350,000 mortgage:**

```{r}
options(max.print = 999999)

periods <- 1:6
df <- data.frame(
  Period       = periods,
  Payment      = round(pmt(0.0325/12, 180, 350000, 0, FALSE), 2),
  Interest     = round(sapply(periods, function(p) ipmt(0.0325/12, p, 180, 350000, 0, FALSE)), 2),
  CumInterest  = round(sapply(periods, function(p) cumipmt(0.0325/12, 180, 350000, 1, p, FALSE)), 2),
  Principal    = round(sapply(periods, function(p) ppmt(0.0325/12, p, 180, 350000, 0, FALSE)), 2),
  CumPrincipal = round(sapply(periods, function(p) cumprinc(0.0325/12, 180, 350000, 1, p, FALSE)), 2)
)

df$Balance <- round(350000 + df$CumPrincipal, 2)
df
```

---

<div class="fn-card">cumipmt(rate, nper, pv, start_period, end_period, pmt_at_beginning)</div>
<div class="purpose">Cumulative Interest — the total interest paid between two
specific periods. Used for annual mortgage interest statements, tax filings,
and comparing loan products. Returns NA if start_period > end_period or
either is outside [1, nper].</div>

```{r}
# Total interest in the first month of a $125,000 mortgage at 9% over 30 years
# Excel: CUMIPMT(0.09/12, 360, 125000, 1, 1, FALSE) = -937.50
cumipmt(0.09/12, 360, 125000, 1, 1, FALSE)

# Guard: start > end returns NA
cumipmt(0.09/12, 360, 125000, 5, 1, FALSE)

# Guard: end > nper returns NA
cumipmt(0.09/12, 12, 125000, 1, 999, FALSE)
```

---

<div class="fn-card">cumprinc(rate, nper, pv, start_period, end_period, pmt_at_beginning)</div>
<div class="purpose">Cumulative Principal — the total principal repaid between
two periods. Paired with cumipmt() to show exactly how much a loan balance
has been reduced over any time span.</div>

```{r}
# Principal repaid in the first month of the same mortgage
# Excel: CUMPRINC(0.09/12, 360, 125000, 1, 1, FALSE) = -68.28
cumprinc(0.09/12, 360, 125000, 1, 1, FALSE)

# Guard: start > end returns NA
cumprinc(0.09/12, 360, 125000, 10, 1, FALSE)
```

---

<div class="fn-card">nper(rate, pmt, pv, fv, pmt_at_beginning)</div>
<div class="purpose">Number of Periods — how long will it take to pay off a
loan or reach a savings goal? Answers "how many months until my mortgage is
paid off?" or "how many years until my portfolio reaches $1 million?"</div>

```{r}
# How many months to grow -$1,000 to $10,000 at 12% with $100/month?
# Excel: NPER(0.12/12, -100, -1000, 10000, TRUE) = 59.67
nper(0.12/12, -100, -1000, 10000, TRUE)
```

---

<div class="fn-card">rate(nper, pmt, pv, fv, pmt_at_beginning, guess)</div>
<div class="purpose">Interest Rate — what rate is implied by a known set of
loan or savings terms? Uses Newton-Raphson iteration (up to 300 iterations,
tolerance 1e-10). Returns NA if the solver does not converge — try a different
initial guess. Returns NA if nper = 0.</div>

```{r}
# Annual rate implied by 48 monthly payments of $200 on an $8,000 loan
# Excel: RATE(48, -200, 8000, 0, FALSE, 0.1) * 12 = 9.24%
rate(48, -200, 8000, 0, FALSE, 0.1) * 12

# Guard: nper = 0 returns NA
rate(0, -200, 8000, 0, FALSE, 0.1)
```

---

<div class="fn-card">ispmt(rate, per, nper, pv)</div>
<div class="purpose">Interest Paid (Straight-Line Loan) — calculates interest
paid in a specific period under a straight-line repayment schedule (equal
principal instalments). Different from ipmt(), which applies to standard
annuity loans with equal total payments. Returns NA if per is outside
[1, nper].</div>

```{r}
# Interest in period 1 of an 8,000,000 loan at 10% over 36 periods
# Excel: ISPMT(0.10/12, 1, 36, 8000000) = -66666.67
ispmt(0.10/12, 1, 36, 8000000)

# Guard: per = 0 returns NA
ispmt(0.05, 0, 12, 10000)

# Guard: per > nper returns NA
ispmt(0.05, 99, 12, 10000)
```

---

## 📈 Cash Flow Functions

Cash flow functions evaluate investment decisions. They answer questions like
"does this project create value?" and "what return does this investment produce?"
They are the primary tools of capital budgeting, project appraisal, and
private equity valuation.

---

<div class="fn-card">npv(rate, values)</div>
<div class="purpose">Net Present Value — the present value of all future cash
flows discounted at a given rate. A positive NPV means the investment creates
value at that discount rate. Note: like Excel, npv() treats the first element
as a period-1 cash flow — add any time-0 investment separately.</div>

```{r}
# Is a $10,000 investment worth it given these annual cash flows at 10%?
# Excel: NPV(0.10, -10000, 3000, 4200, 6800) = 1188.44
npv(0.10, c(-10000, 3000, 4200, 6800))

# With period-0 investment added separately
npv(0.08, c(8000, 9200, 10000, 12000, 14500)) + (-40000)

# Guard: empty vector returns NA
npv(0.10, numeric(0))
```

---

<div class="fn-card">irr(values)</div>
<div class="purpose">Internal Rate of Return — the discount rate at which the
NPV of a series of cash flows equals zero. A project whose IRR exceeds the
cost of capital creates value. Requires at least one positive and one negative
cash flow. Uses bisection — robust against non-standard cash flow patterns.
Returns NA if no solution exists.</div>

```{r}
# IRR of a 5-year project — Excel: IRR(-70000,...,26000) = 8.66%
irr(c(-70000, 12000, 15000, 18000, 21000, 26000))

# IRR after 4 years only (still negative — hasn't broken even yet)
irr(c(-70000, 12000, 15000, 18000, 21000))

# IRR after 2 years
irr(c(-70000, 12000, 15000))

# Guard: all positive (no valid IRR) returns NA
irr(c(100, 200, 300))

# Guard: all negative returns NA
irr(c(-100, -200, -300))

# Guard: single value returns NA
irr(c(-1000))
```

---

<div class="fn-card">mirr(values, finance_rate, reinvest_rate)</div>
<div class="purpose">Modified IRR — addresses IRR's reinvestment assumption
by using separate rates for financing negative flows (cost of borrowing) and
reinvesting positive flows (reinvestment return). More realistic than IRR
for most real-world projects. Returns NA if all cash flows have the same sign.</div>

```{r}
# 5-year project, 10% financing, 12% reinvestment
# Excel: MIRR(-120000,...,46000, 0.10, 0.12) = 12.61%
mirr(c(-120000, 39000, 30000, 21000, 37000, 46000), 0.10, 0.12)

# After 3 years only (negative — not yet viable)
mirr(c(-120000, 39000, 30000, 21000), 0.10, 0.12)

# With 14% reinvestment rate
mirr(c(-120000, 39000, 30000, 21000, 37000, 46000), 0.10, 0.14)

# Guard: all positive returns NA
mirr(c(100, 200, 300), 0.10, 0.12)
```

---

<div class="fn-card">xnpv(rate, values, dates)</div>
<div class="purpose">NPV with Irregular Dates — like npv() but discounts each
cash flow by its exact actual date rather than assuming equal periods. Essential
for real-world projects where cash flows don't fall on regular calendar dates.
Dates must be "YYYY-MM-DD" strings, the same length as values.</div>

```{r}
# Irregular cash flows at actual dates, discounted at 9%
# Excel: XNPV(0.09, values, dates) = 2086.65
xnpv(0.09,
  c(-10000, 2750, 4250, 3250, 2750),
  c("2008-01-01","2008-03-01","2008-10-30","2009-02-15","2009-04-01"))

# Guard: mismatched lengths return NA
xnpv(0.09, c(-10000, 2750), "2008-01-01")

# Guard: bad date string returns NA
xnpv(0.09, c(-10000, 2750), c("not-a-date","2008-03-01"))
```

---

<div class="fn-card">xirr(values, dates)</div>
<div class="purpose">IRR with Irregular Dates — the date-based counterpart to
irr(). Finds the discount rate that makes XNPV equal zero. The standard tool
for private equity, real estate, and any investment with irregularly timed
cash flows. Returns NA if no solution exists.</div>

```{r}
# Excel: XIRR(values, dates) = 37.34%
xirr(c(-10000, 2750, 4250, 3250, 2750),
     c("2008-01-01","2008-03-01","2008-10-30","2009-02-15","2009-04-01"))

# Guard: all positive returns NA
xirr(c(100, 200, 300), c("2020-01-01","2020-06-01","2021-01-01"))

# Guard: mismatched lengths return NA
xirr(c(-10000, 2750), "2008-01-01")
```

---

## 🏭 Depreciation Functions

Depreciation functions allocate the cost of an asset over its useful life.
Different methods produce different profiles — straight-line for simplicity,
accelerated methods for faster early write-downs (tax advantages), and French
methods (AMORLINC, AMORDEGRC) required by European accounting standards.

---

<div class="fn-card">sln(cost, salvage, life)</div>
<div class="purpose">Straight-Line Depreciation — allocates an equal charge
each period. The simplest and most transparent method: (cost - salvage) / life.
Returns NA if life = 0.</div>

```{r}
# $30,000 asset, $7,500 salvage, 10-year life
# Excel: SLN(30000, 7500, 10) = 2250
sln(30000, 7500, 10)

# Guard: life = 0 returns NA
sln(30000, 7500, 0)
```

---

<div class="fn-card">syd(cost, salvage, life, per)</div>
<div class="purpose">Sum-of-Years-Digits Depreciation — an accelerated method
that front-loads depreciation. Year 1 gets the largest charge; it decreases
each year. Useful for assets that lose value quickly in early life (vehicles,
technology equipment).</div>

```{r}
# Year 1: $30,000 asset, $7,500 salvage, 10-year life
# Excel: SYD(30000, 7500, 10, 1) = 4090.91
syd(30000, 7500, 10, 1)
```

---

<div class="fn-card">db(cost, salvage, life, per, month)</div>
<div class="purpose">Fixed-Declining Balance Depreciation — applies a fixed
percentage rate to the declining book value each period. The rate is computed
from cost, salvage, and life. The month argument handles partial first years
when an asset is acquired mid-year.</div>

```{r}
# $1,000,000 asset, $100,000 salvage, 6-year life, acquired in month 6
# Excel: DB(1000000, 100000, 6, 1, 7) = 186083.33
db(1000000, 100000, 6, 1, 7)
```

---

<div class="fn-card">ddb(cost, salvage, life, per, factor)</div>
<div class="purpose">Double-Declining Balance Depreciation — applies double
the straight-line rate to the declining book value. The most aggressive
standard accelerated method. The factor parameter controls the multiplier
(2.0 = double-declining, the Excel default).</div>

```{r}
# Year 1: $2,400 asset, $300 salvage, 10-year life
# Excel: DDB(2400, 300, 10, 1, 2) = 480
ddb(2400, 300, 10, 1, 2)
```

---

<div class="fn-card">vdb(cost, salvage, life, start_period, end_period, factor, no_switch)</div>
<div class="purpose">Variable Declining Balance — the most flexible
depreciation function. Computes depreciation over any time span including
fractional periods, and optionally switches to straight-line when it becomes
more advantageous. Ideal for pro-rated partial-year calculations.</div>

```{r}
# First month of a $2,400 asset with life expressed in months
# Excel: VDB(2400, 300, 10*12, 0, 1, 2, FALSE) = 40
vdb(2400, 300, 10*12, 0, 1, 2, FALSE)

# Full first year
vdb(2400, 300, 10, 0, 1, 2, FALSE)
```

**Depreciation comparison over 10 years:**

```{r}
years <- 1:10
data.frame(
  Year         = years,
  StraightLine = round(sln(30000, 7500, 10), 0),
  SumOfYears   = round(sapply(years, function(p) syd(30000, 7500, 10, p)), 0),
  DblDeclining = round(sapply(years, function(p) ddb(30000, 7500, 10, p, 2)), 0)
)
```

---

<div class="fn-card">amorlinc(cost, date_purchased, first_period, salvage, period, rate, basis)</div>
<div class="purpose">French Straight-Line Depreciation — a European accounting
standard where the first period's charge is pro-rated by the actual number of
days between purchase and period end. Dates are Excel serial numbers
(days since 1899-12-30).</div>

```{r}
# Excel date 39679 = 2008-09-23, first period end 39813 = 2009-02-01
# Excel: AMORLINC(2400, 39679, 39813, 300, 1, 0.15, 1) = 360
amorlinc(2400, 39679, 39813, 300, 1, 0.15, 1)
```

---

<div class="fn-card">amordegrc(cost, date_purchased, first_period, salvage, period, rate, basis)</div>
<div class="purpose">French Degressive Depreciation — like AMORLINC but applies
a coefficient based on the asset's useful life that accelerates the write-down.
Required by French tax law and some other European accounting standards.
Returns NA if rate = 0.</div>

```{r}
# Same asset — degressive method gives a higher first-year charge
# Excel: AMORDEGRC(2400, 39679, 39813, 300, 1, 0.15, 1) = 776
amordegrc(2400, 39679, 39813, 300, 1, 0.15, 1)

# Guard: rate = 0 returns NA
amordegrc(2400, 39679, 39813, 300, 1, 0.0, 1)
```

---

## 🗓️ Coupon Date Functions

Coupon date functions calculate dates and day counts for bond coupon periods.
They are essential for accrued interest calculations, settlement processing,
and bond analytics. All date arguments use `"YYYY-MM-DD"` format.

**Day-count basis codes:**

| Code | Convention | Common Use |
|------|-----------|------------|
| 0 | US 30/360 | US corporate and agency bonds |
| 1 | Actual/Actual | US Treasuries, government bonds |
| 2 | Actual/360 | US money markets |
| 3 | Actual/365 | UK Gilts |
| 4 | European 30/360 | Eurobonds |

All coupon functions return NA if settlement >= maturity (prevents infinite loops).

---

<div class="fn-card">coupdaybs(settlement, maturity, frequency, basis)</div>
<div class="purpose">Days from Coupon Start to Settlement — the number of days
elapsed since the beginning of the current coupon period (i.e., since the last
coupon date). Used to compute the accrued interest the bond buyer owes the
seller.</div>

```{r}
# Settled 2011-01-25, matures 2011-11-15, semiannual, actual/actual
# Excel: COUPDAYBS("2011-01-25","2011-11-15",2,1) = 71
coupdaybs("2011-01-25", "2011-11-15", 2, 1)

# Guard: settlement > maturity returns NA
coupdaybs("2025-01-01", "2020-01-01", 2, 1)

# Guard: settlement = maturity returns NA
coupdaybs("2020-01-01", "2020-01-01", 2, 1)
```

---

<div class="fn-card">coupdays(settlement, maturity, frequency, basis)</div>
<div class="purpose">Total Days in Coupon Period — the full length of the
coupon period containing the settlement date. Combined with coupdaybs(), this
gives the fraction of the period elapsed: coupdaybs / coupdays = accrual
fraction.</div>

```{r}
# Excel: COUPDAYS("2011-01-25","2011-11-15",2,1) = 181
coupdays("2011-01-25", "2011-11-15", 2, 1)
```

---

<div class="fn-card">coupdaysnc(settlement, maturity, frequency, basis)</div>
<div class="purpose">Days from Settlement to Next Coupon — the number of days
remaining until the next coupon payment. Equals coupdays - coupdaybs. Used in
bond pricing to determine the timing of the nearest cash flow.</div>

```{r}
# Excel: COUPDAYSNC("2011-01-25","2011-11-15",2,1) = 110
coupdaysnc("2011-01-25", "2011-11-15", 2, 1)
```

---

<div class="fn-card">coupncd(settlement, maturity, frequency, basis)</div>
<div class="purpose">Next Coupon Date — the date of the next coupon payment,
returned as an Excel serial number. Convert to an R Date using
as.Date(result, origin = "1899-12-30").</div>

```{r}
serial <- coupncd("2011-01-25", "2011-11-15", 2, 1)
as.Date(serial, origin = "1899-12-30")
```

---

<div class="fn-card">couppcd(settlement, maturity, frequency, basis)</div>
<div class="purpose">Previous Coupon Date — the date of the most recent coupon
payment before the settlement date. This is the start of the current accrual
period, used to compute accrued interest.</div>

```{r}
serial <- couppcd("2011-01-25", "2011-11-15", 2, 1)
as.Date(serial, origin = "1899-12-30")
```

---

<div class="fn-card">coupnum(settlement, maturity, frequency, basis)</div>
<div class="purpose">Number of Coupons Remaining — how many coupon payments
remain between settlement and maturity. Determines the number of discounting
periods in bond pricing.</div>

```{r}
# Excel: COUPNUM("2007-01-25","2008-11-15",2,1) = 4
coupnum("2007-01-25", "2008-11-15", 2, 1)
```

---

## 🏦 Bond & Security Functions

Bond functions implement the standard market conventions used by Bloomberg,
Refinitiv, and fixed-income trading desks worldwide. They price bonds, compute
yields, measure interest-rate sensitivity, and calculate accrued interest.

---

<div class="fn-card">price(settlement, maturity, rate, yld, redemption, frequency, basis)</div>
<div class="purpose">Clean Bond Price — the present value of all future coupon
and principal cash flows, expressed per $100 face value. Returns the clean
price (excluding accrued interest), which is the standard market quotation.
Returns NA if settlement >= maturity.</div>

```{r}
# 5.75% coupon bond, 6.5% yield, semiannual, 30/360
# Excel: PRICE("2008-02-15","2017-11-15",0.0575,0.065,100,2,0) = 94.63
price("2008-02-15", "2017-11-15", 0.0575, 0.065, 100, 2, 0)

# Guard: settlement > maturity returns NA
price("2020-01-01", "2010-01-01", 0.05, 0.06, 100, 2, 0)

# Guard: bad date returns NA
price("bad-date", "2017-11-15", 0.0575, 0.065, 100, 2, 0)
```

---

<div class="fn-card">yield_(settlement, maturity, rate, pr, redemption, frequency, basis)</div>
<div class="purpose">Yield to Maturity — the internal rate of return of a bond
given its clean market price. The inverse of price(). Note the trailing
underscore — this avoids a conflict with R's built-in yield keyword.</div>

```{r}
# YTM of the same bond priced at 95.04287
# Excel: YIELD("2008-02-15","2016-11-15",0.0575,95.04287,100,2,0) = 6.5%
yield_("2008-02-15", "2016-11-15", 0.0575, 95.04287, 100, 2, 0)
```

---

<div class="fn-card">pricedisc(settlement, maturity, discount, redemption, basis)</div>
<div class="purpose">Price of a Discounted Security — calculates the price of
a zero-coupon or discount security (Treasury bill, commercial paper) from its
annualised discount rate.</div>

```{r}
# Excel: PRICEDISC("2008-02-16","2008-03-01",0.0525,100,2) = 99.795
pricedisc("2008-02-16", "2008-03-01", 0.0525, 100, 2)
```

---

<div class="fn-card">yielddisc(settlement, maturity, pr, redemption, basis)</div>
<div class="purpose">Yield of a Discounted Security — the annualised yield of
a discount security from its price. The inverse of pricedisc().</div>

```{r}
# Excel: YIELDDISC("2008-02-16","2008-03-01",99.795,100,2) = 5.25%
yielddisc("2008-02-16", "2008-03-01", 99.795, 100, 2)
```

---

<div class="fn-card">pricemat(settlement, maturity, issue, rate, yld, basis)</div>
<div class="purpose">Price of Security Paying at Maturity — prices a security
that pays all interest at maturity rather than periodically (term deposit,
zero-coupon bond).</div>

```{r}
pricemat("2008-02-15", "2008-04-13", "2007-11-11", 0.061, 0.061, 0)
```

---

<div class="fn-card">yieldmat(settlement, maturity, issue, rate, pr, basis)</div>
<div class="purpose">Yield of Security Paying at Maturity — the yield of a
security that pays interest at maturity. The inverse of pricemat().</div>

```{r}
yieldmat("2008-03-15", "2008-11-03", "2007-11-08", 0.0625, 100.0123, 0)
```

---

<div class="fn-card">disc(settlement, maturity, pr, redemption, basis)</div>
<div class="purpose">Discount Rate — the annualised discount rate of a
security from its price and redemption value. Used for Treasury bills
and other discount instruments.</div>

```{r}
disc("2008-02-16", "2008-03-01", 97.975, 100, 2)
```

---

<div class="fn-card">intrate(settlement, maturity, investment, redemption, basis)</div>
<div class="purpose">Interest Rate for Fully Invested Security — the implied
annual interest rate for a security where the investor pays a known price and
receives a known redemption value, with no intermediate coupon payments.</div>

```{r}
# Excel: INTRATE("2008-02-15","2008-05-15",1000000,1014420,2) = 5.768%
intrate("2008-02-15", "2008-05-15", 1000000, 1014420, 2)
```

---

<div class="fn-card">received(settlement, maturity, investment, discount, basis)</div>
<div class="purpose">Amount Received at Maturity — the total amount an investor
receives at maturity given an initial investment and annual discount rate.
The inverse of intrate().</div>

```{r}
# Excel: RECEIVED("2008-02-15","2008-05-15",1000000,0.0575,2) = 1014584.65
received("2008-02-15", "2008-05-15", 1000000, 0.0575, 2)
```

---

<div class="fn-card">duration(settlement, maturity, coupon, yld, frequency, basis)</div>
<div class="purpose">Macaulay Duration — the weighted average time (in years)
to receive all the bond's cash flows, weighted by their present values. A
measure of interest-rate risk: bonds with longer duration experience larger
price swings when yields change.</div>

```{r}
# 8% coupon bond, 9% yield, semiannual, actual/actual
# Excel: DURATION("2008-01-01","2016-01-01",0.08,0.09,2,1) = 5.9938
duration("2008-01-01", "2016-01-01", 0.08, 0.09, 2, 1)
```

---

<div class="fn-card">mduration(settlement, maturity, coupon, yld, frequency, basis)</div>
<div class="purpose">Modified Duration — Macaulay Duration divided by
(1 + yield/frequency). Measures the percentage change in a bond's price for
a 1 percentage point (100 basis point) change in yield. The standard tool
for interest-rate risk management and hedging.</div>

```{r}
# Excel: MDURATION("2008-01-01","2016-01-01",0.08,0.09,2,1) = 5.7355
mduration("2008-01-01", "2016-01-01", 0.08, 0.09, 2, 1)
```

---

<div class="fn-card">accrint(issue, first_interest, settlement, rate, par, frequency, basis)</div>
<div class="purpose">Accrued Interest (Periodic Coupon) — the interest accrued
on a bond from its issue date to the settlement date. The bond buyer compensates
the seller for this amount in addition to the clean price, making up the
"dirty price" paid at settlement.</div>

```{r}
accrint("2008-03-01", "2008-08-31", "2008-05-01", 0.10, 1000, 2, 0)
```

---

<div class="fn-card">accrintm(issue, settlement, rate, par, basis)</div>
<div class="purpose">Accrued Interest at Maturity — for securities that pay
all interest at maturity rather than periodically, this computes the total
interest accrued from issue to settlement date.</div>

```{r}
# Excel: ACCRINTM("2008-04-01","2008-06-15",0.10,1000,3) = 20.54
accrintm("2008-04-01", "2008-06-15", 0.10, 1000, 3)
```

---

<div class="fn-card">oddfprice(settlement, maturity, issue, first_coupon, rate, yld, redemption, frequency, basis)</div>
<div class="purpose">Price with Odd First Period — prices a bond where the
first coupon period is shorter or longer than a normal period. Common with
new bond issues where the first coupon covers a partial period from the issue
date to the first scheduled coupon date.</div>

```{r}
# Excel: ODDFPRICE(...) = 113.60
oddfprice("2008-11-11","2021-03-01","2008-10-15","2009-03-01",
          0.0785, 0.0625, 100, 2, 1)
```

---

<div class="fn-card">oddfyield(settlement, maturity, issue, first_coupon, rate, pr, redemption, frequency, basis)</div>
<div class="purpose">Yield with Odd First Period — the yield of a bond with an
odd first coupon period given its clean price. The inverse of oddfprice().</div>

```{r}
oddfyield("2008-11-11","2021-03-01","2008-10-15","2009-03-01",
          0.0785, 113.60, 100, 2, 1)
```

---

<div class="fn-card">oddlprice(settlement, maturity, last_interest, rate, yld, redemption, frequency, basis)</div>
<div class="purpose">Price with Odd Last Period — prices a bond where the
final coupon period is shorter than a normal period (the bond matures between
two scheduled coupon dates).</div>

```{r}
# Excel: ODDLPRICE(...) = 99.88
oddlprice("2008-02-07","2008-06-15","2007-10-15",
          0.0375, 0.0405, 100, 2, 0)
```

---

<div class="fn-card">oddlyield(settlement, maturity, last_interest, rate, pr, redemption, frequency, basis)</div>
<div class="purpose">Yield with Odd Last Period — the yield of a bond with
an odd last coupon period given its price. The inverse of oddlprice().</div>

```{r}
# Excel: ODDLYIELD(...) = 4.52%
oddlyield("2008-04-20","2008-06-15","2007-12-24",
          0.0375, 99.875, 100, 2, 0)
```

---

## 🔧 Miscellaneous Functions

This group covers interest rate conversions, Treasury bill analytics, dollar
price notation, and investment growth utilities.

---

<div class="fn-card">effect(nominal_rate, npery)</div>
<div class="purpose">Effective Annual Rate — converts a nominal (stated) annual
rate to the true effective annual rate, accounting for compounding within the
year. A 5.25% rate compounded quarterly has a higher effective return than
5.25% compounded annually. Returns NA if npery = 0.</div>

```{r}
# Excel: EFFECT(0.0525, 4) = 5.3543%
effect(0.0525, 4)

# Guard: npery = 0 returns NA
effect(0.05, 0)
```

---

<div class="fn-card">nominal(effect_rate, npery)</div>
<div class="purpose">Nominal Annual Rate — the inverse of effect(). Converts
a known effective annual rate back to its nominal equivalent for a given
compounding frequency. Used to compare products with different compounding
periods on a level playing field.</div>

```{r}
# Excel: NOMINAL(0.053543, 4) = 5.25% — confirms effect() is invertible
nominal(0.053543, 4)
```

---

<div class="fn-card">dollarde(fractional_dollar, fraction)</div>
<div class="purpose">Dollar Price: Fraction to Decimal — converts a price
quoted in fractional notation (standard in US Treasury markets) to a decimal.
For example, "101-02" means 101 and 2/32nds = 101.0625. Returns NA if
fraction = 0.</div>

```{r}
# 1 and 2/16ths = 1.125
# Excel: DOLLARDE(1.02, 16) = 1.125
dollarde(1.02, 16)
```

---

<div class="fn-card">dollarfr(decimal_dollar, fraction)</div>
<div class="purpose">Dollar Price: Decimal to Fraction — converts a decimal
price back to fractional notation. The inverse of dollarde(). Used to present
bond prices in the conventional Treasury market format.</div>

```{r}
# 1.125 in 16ths = 1 + 2/16, displayed as 1.02
# Excel: DOLLARFR(1.125, 16) = 1.02
dollarfr(1.125, 16)
```

---

<div class="fn-card">fvschedule(principal, schedule)</div>
<div class="purpose">Future Value with Variable Rates — grows a principal by
applying a different interest rate each period. Used for variable-rate
mortgages, bonds with step-up coupons, or any scenario where the rate
changes over time. Returns NA if the schedule is empty.</div>

```{r}
# Excel: FVSCHEDULE(1, {0.09, 0.11, 0.10}) = 1.3309
fvschedule(1, c(0.09, 0.11, 0.10))

# Guard: empty schedule returns NA
fvschedule(1, numeric(0))
```

---

<div class="fn-card">rri(nper, pv, fv)</div>
<div class="purpose">Equivalent Interest Rate — the constant periodic rate
required to grow a present value to a future value over a given number of
periods. A simple way to compute Compound Annual Growth Rate (CAGR).
Returns NA if nper = 0 or pv = 0.</div>

```{r}
# CAGR: $10,000 to $11,000 over 96 months
rri(96, 10000, 11000)

# Guard: pv = 0 returns NA
rri(10, 0, 1000)

# Guard: nper = 0 returns NA
rri(0, 1000, 2000)
```

---

<div class="fn-card">pduration(rate, pv, fv)</div>
<div class="purpose">Periods to Target Value — how many periods at a given
interest rate does it take to grow a present value to a target? The time
dimension of a growth calculation — complementary to rri() which solves
for rate.</div>

```{r}
# Excel: PDURATION(0.025, 2000, 2200) = 3.8599
pduration(0.025, 2000, 2200)
```

---

<div class="fn-card">tbillprice(settlement, maturity, discount)</div>
<div class="purpose">Treasury Bill Price — the price per $100 face value of
a US Treasury bill from its annualised discount rate. Uses the 360-day
convention standard in US money markets.</div>

```{r}
# Excel: TBILLPRICE("2008-03-31","2008-06-01",0.09) = 98.45
tbillprice("2008-03-31", "2008-06-01", 0.09)
```

---

<div class="fn-card">tbillyield(settlement, maturity, pr)</div>
<div class="purpose">Treasury Bill Yield — the annualised discount yield of a
T-bill from its price. Uses the 360-day money market convention. The inverse
of tbillprice().</div>

```{r}
# Excel: TBILLYIELD("2008-03-31","2008-06-01",98.45) = 9.14%
tbillyield("2008-03-31", "2008-06-01", 98.45)
```

---

<div class="fn-card">tbilleq(settlement, maturity, discount)</div>
<div class="purpose">T-Bill Bond-Equivalent Yield — converts a T-bill's
discount rate to a bond-equivalent yield (BEY), enabling fair comparison
with coupon-bearing bonds that use a 365-day year. The formula differs for
bills under and over 182 days. Returns NA on invalid dates.</div>

```{r}
# Excel: TBILLEQ("2008-03-31","2008-06-01",0.0914) = 9.42%
tbilleq("2008-03-31", "2008-06-01", 0.0914)

# Guard: bad date returns NA
tbilleq("bad-date", "2008-06-01", 0.09)
```

---

## Using with dplyr

All functions work naturally inside `dplyr::mutate()`:

```{r eval=FALSE}
library(dplyr)

loans <- data.frame(
  borrower = c("Alice",  "Bob",    "Carol"),
  amount   = c(250000,   180000,   320000),
  rate_pa  = c(0.035,    0.042,    0.038),
  years    = c(20,       15,       25)
)

loans |>
  mutate(
    monthly_payment = round(pmt(rate_pa/12, years*12, amount, 0, FALSE), 2),
    total_interest  = round(-cumipmt(rate_pa/12, years*12, amount,
                                    1, years*12, FALSE), 0),
    effective_rate  = round(effect(rate_pa, 12) * 100, 3)
  )
```

---

## Key Differences from Excel

| Excel | R | Note |
|-------|---|------|
| `PMT(rate, nper, pv)` | `pmt(rate, nper, pv, 0, FALSE)` | All args required in R |
| `YIELD(...)` | `yield_(...)` | Trailing `_` avoids R keyword |
| `IRR(A1:A6)` | `irr(c(-70000, 12000, ...))` | R vector, not cell range |
| `XIRR(vals, dates)` | `xirr(vals, c("2024-01-01",...))` | Dates as `"YYYY-MM-DD"` |
| `#NUM!` error | `NA` | Graceful failure, never crashes |

---

## Getting Help

```{r eval=FALSE}
?pmt       # Annuity functions
?irr       # Cash flow functions
?price     # Bond functions
?duration  # Duration / risk
```

Full function reference: [tor-storli.github.io/FinancialFunctionsR](https://tor-storli.github.io/FinancialFunctionsR/)
