---
title: "Getting Started with FinancialFunctions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with FinancialFunctions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  digits    = 4
)
library(FinancialFunctions)
```

## Overview

**FinancialFunctions** provides 55 Excel-compatible financial functions implemented
in Rust for high performance. All functions:

- Match Excel output exactly
- Return `NA` (never crash) on invalid input
- Work on scalar values and inside `dplyr` pipelines

```{r}
library(FinancialFunctions)
```

---

## 💰 Annuity Functions

Annuity functions calculate the components of loan repayments and savings plans.

### PMT — Periodic Payment

What is the monthly payment on a mortgage?

```{r}
# $350,000 mortgage at 3.25% annual rate over 15 years
pmt(rate = 0.0325/12, nper = 180, pv = 350000, fv = 0, pmt_at_beginning = FALSE)
```

### FV — Future Value

How much will monthly savings of $500 grow to over 20 years at 6%?

```{r}
fv(rate = 0.06/12, nper = 240, pmt = -500, pv = 0, pmt_at_beginning = FALSE)
```

### PV — Present Value

What is a pension worth today if it pays $2,000/month for 20 years at 5%?

```{r}
pv(rate = 0.05/12, nper = 240, pmt = 2000, fv = 0, pmt_at_beginning = FALSE)
```

### RATE — Interest Rate

What annual rate is implied by a $350,000 loan with $2,459 monthly payments over 15 years?

```{r}
rate(nper = 180, pmt = -2459, pv = 350000, fv = 0,
     pmt_at_beginning = FALSE, guess = 0.05) * 12
```

### NPER — Number of Periods

How many months to pay off a $10,000 loan at 8% with $200 monthly payments?

```{r}
nper(rate = 0.08/12, pmt = -200, pv = 10000, fv = 0, pmt_at_beginning = FALSE)
```

### Loan Amortisation Table

Build a full amortisation schedule using `ipmt` and `ppmt`:

```{r}
periods <- 1:12
schedule <- data.frame(
  Period    = periods,
  Payment   = round(pmt(0.04/12, 360, 300000, 0, FALSE), 2),
  Interest  = round(sapply(periods, function(p) ipmt(0.04/12, p, 360, 300000, 0, FALSE)), 2),
  Principal = round(sapply(periods, function(p) ppmt(0.04/12, p, 360, 300000, 0, FALSE)), 2)
)
head(schedule, 6)
```

### Cumulative Interest and Principal

How much interest is paid in the first year of a $300,000 mortgage at 4%?

```{r}
# Cumulative interest for periods 1 through 12
cumipmt(rate = 0.04/12, nper = 360, pv = 300000,
        start_period = 1, end_period = 12, pmt_at_beginning = FALSE)

# Cumulative principal for the same period
cumprinc(rate = 0.04/12, nper = 360, pv = 300000,
         start_period = 1, end_period = 12, pmt_at_beginning = FALSE)
```

---

## 📈 Cash Flow Functions

### NPV — Net Present Value

Is a $40,000 investment worth it given these projected cash flows at 8% discount?

```{r}
# Initial investment at time 0 + NPV of future flows
npv(rate = 0.08, values = c(8000, 9200, 10000, 12000, 14500)) + (-40000)
```

A positive NPV means the investment creates value.

### IRR — Internal Rate of Return

What is the return on a 5-year project?

```{r}
irr(c(-70000, 12000, 15000, 18000, 21000, 26000))
```

```{r}
# Convert to percentage
irr(c(-70000, 12000, 15000, 18000, 21000, 26000)) * 100
```

### MIRR — Modified IRR

MIRR uses separate rates for financing (cost of capital) and reinvestment:

```{r}
mirr(
  values       = c(-120000, 39000, 30000, 21000, 37000, 46000),
  finance_rate  = 0.10,
  reinvest_rate = 0.12
)
```

### XNPV and XIRR — Irregular Cash Flows

When cash flows don't fall on regular intervals, use the date-based versions:

```{r}
values <- c(-10000, 2750, 4250, 3250, 2750)
dates  <- c("2024-01-01", "2024-03-01", "2024-10-30", "2025-02-15", "2025-04-01")

xnpv(rate = 0.09, values = values, dates = dates)
```

```{r}
xirr(values = values, dates = dates)
```

---

## 🏭 Depreciation Functions

### SLN — Straight-Line Depreciation

```{r}
# $30,000 asset, $7,500 salvage value, 10-year life
sln(cost = 30000, salvage = 7500, life = 10)
```

### DDB — Double-Declining Balance

```{r}
# Year 1 depreciation on a $2,400 asset with 10-year life
ddb(cost = 2400, salvage = 300, life = 10, per = 1, factor = 2)
```

### Depreciation Schedule

Compare methods over the full asset life:

```{r}
asset_life <- 1:10
data.frame(
  Year         = asset_life,
  StraightLine = round(sln(30000, 7500, 10), 0),
  SumOfYears   = round(sapply(asset_life, function(p) syd(30000, 7500, 10, p)), 0),
  DeclBalance  = round(sapply(asset_life, function(p) ddb(30000, 7500, 10, p, 2)), 0)
)
```

---

## 🗓️ Coupon Date Functions

These functions calculate dates and day counts for bond coupon periods.

```{r}
settle  <- "2024-01-25"
mature  <- "2030-11-15"
freq    <- 2        # semiannual
basis   <- 1        # actual/actual

# Days from last coupon to settlement
coupdaybs(settle, mature, freq, basis)

# Total days in coupon period
coupdays(settle, mature, freq, basis)

# Days from settlement to next coupon
coupdaysnc(settle, mature, freq, basis)

# Number of coupons remaining
coupnum(settle, mature, freq, basis)
```

---

## 🏦 Bond & Security Functions

### PRICE — Clean Bond Price

What is the clean price of a 5.75% coupon bond maturing in 2030?

```{r}
price(
  settlement = "2024-02-15",
  maturity   = "2030-11-15",
  rate       = 0.0575,
  yld        = 0.065,
  redemption = 100,
  frequency  = 2,
  basis      = 0
)
```

### YIELD — Bond Yield

What yield does a bond with price 94.63 imply?

```{r}
yield_(
  settlement = "2024-02-15",
  maturity   = "2030-11-15",
  rate       = 0.0575,
  pr         = 94.63,
  redemption = 100,
  frequency  = 2,
  basis      = 0
)
```

### DURATION and MDURATION

```{r}
# Macaulay Duration
duration(
  settlement = "2024-01-01",
  maturity   = "2032-01-01",
  coupon     = 0.08,
  yld        = 0.09,
  frequency  = 2,
  basis      = 1
)

# Modified Duration (price sensitivity to yield changes)
mduration(
  settlement = "2024-01-01",
  maturity   = "2032-01-01",
  coupon     = 0.08,
  yld        = 0.09,
  frequency  = 2,
  basis      = 1
)
```

### Treasury Bills

```{r}
# T-bill price
tbillprice(settlement = "2024-03-31", maturity = "2024-06-01", discount = 0.09)

# T-bill yield
tbillyield(settlement = "2024-03-31", maturity = "2024-06-01", pr = 98.45)

# Bond-equivalent yield
tbilleq(settlement = "2024-03-31", maturity = "2024-06-01", discount = 0.0914)
```

---

## 🔧 Miscellaneous Functions

### EFFECT and NOMINAL — Effective vs Nominal Rates

```{r}
# Effective annual rate from 5.25% nominal compounded quarterly
effect(nominal_rate = 0.0525, npery = 4)

# Nominal rate that gives 5.3543% effective when compounded quarterly
nominal(effect_rate = 0.053543, npery = 4)
```

### DOLLARDE and DOLLARFR — Fractional Price Notation

US Treasury prices are quoted in 32nds. Convert to and from decimal:

```{r}
# 101 + 2/32 = 101.0625
dollarde(fractional_dollar = 101.02, fraction = 32)

# 101.0625 back to fractional notation
dollarfr(decimal_dollar = 101.0625, fraction = 32)
```

### FVSCHEDULE — Variable Rate Schedule

How does $10,000 grow over 3 years with different rates each year?

```{r}
fvschedule(principal = 10000, schedule = c(0.05, 0.04, 0.06))
```

### RRI — Equivalent Growth Rate

What annual rate would grow $10,000 to $15,000 over 8 years?

```{r}
rri(nper = 8, pv = 10000, fv = 15000)
```

---

## Using with dplyr

All functions are vectorised and work naturally in `mutate()`:

```{r eval=FALSE}
library(dplyr)

loans <- data.frame(
  borrower = c("Alice", "Bob", "Carol"),
  amount   = c(250000, 180000, 320000),
  rate_pa  = c(0.035, 0.042, 0.038),
  years    = c(20, 15, 25)
)

loans |>
  mutate(
    monthly_payment = round(pmt(rate_pa/12, years*12, amount, 0, FALSE), 2),
    total_cost      = round(monthly_payment * years * 12, 0),
    total_interest  = round(total_cost + amount, 0)
  )
```

---

## Key Differences from Excel

| Excel | R | Note |
|-------|---|------|
| `PMT(rate, nper, pv)` | `pmt(rate, nper, pv, 0, FALSE)` | All args required in R |
| `YIELD(...)` | `yield_(...)` | Trailing underscore avoids R keyword |
| `IRR(A1:A6)` | `irr(c(-70000, 12000, ...))` | Pass R vector |
| `XIRR(vals, dates)` | `xirr(vals, c("2024-01-01", ...))` | Dates as "YYYY-MM-DD" strings |
| `#NUM!` | `NA` | Invalid input returns NA, not an error |

---

## Getting Help

Each function has a help page with parameters and examples:

```{r eval=FALSE}
?pmt
?irr
?price
```

For the full function reference visit the
[package website](https://tor-storli.github.io/FinancialFunctionsR/).
